---
title: "Project 1 - Accessing API's Vignette"
author: "Crista Gregg"
date: "6/10/2021"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# List of Required Packages
```{r message=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)
library(lubridate)
library(knitr)
```

# Functions to access API's
## Franchise Function  

```{r franchise function}
get_franchise <- function(team_id = NA, team_name = NA){
  #connect to API
  get_endpoint <- GET('https://records.nhl.com/site/api/franchise') %>%
    content("text", encoding = 'UTF-8') %>%
    fromJSON(flatten = T) 
  df1 <- get_endpoint$data
  
  #if else statement to get team by name or ID
  ifelse(is.na(team_id) & is.na(team_name), franchise <- df1, 
         ifelse(is.na(team_id) & is.na(team_name) == FALSE, franchise <- filter(df1, fullName == team_name),
                ifelse(is.na(team_id) == FALSE & is.na(team_name), franchise <- filter(df1, id == team_id), 
                       franchise <- filter(df1, id == team_id | fullName == team_name)                                                         ))                                                         )
  return(franchise)
}

get_franchise(3, team_name = 'Montreal Wanderers')
```

## Franchise team totals Function
```{r franchise team totals function}
get_franchise_team_totals <- function(team_id = NA, team_name = NA){
  #connect to API
  get_endpoint <- GET('https://records.nhl.com/site/api/franchise-team-totals') %>%
    content("text", encoding = 'UTF-8') %>%
    fromJSON(flatten = T) 
   df1 <- get_endpoint$data
  
  #if else statement to get team by name or ID
  ifelse(is.na(team_id) & is.na(team_name), franchise_team_totals <- df1, 
         ifelse(is.na(team_id) & is.na(team_name) == FALSE, franchise_team_totals <- filter(df1, teamName == team_name),
                ifelse(is.na(team_id) == FALSE & is.na(team_name), franchise_team_totals <- filter(df1, teamId == team_id), 
                       franchise_team_totals <- filter(df1, teamId == team_id | teamName == team_name)                                                         ))                                                         )
  return(franchise_team_totals)
}

get_franchise_team_totals(2)
```
## Records Function
```{r records function}
#function to call from franchise season records, goalie records, or skater records.
get_franchise_records <- function(record_type , team_id = NA, team_name = NA){
  
  #stop running if incorrect type of record entered
  if (!record_type %in% c('season', 'goalie', 'skater')){
    stop("Please enter record type of 'season', 'goalie', or 'skater'")
  }
  
  #connect to API
  URL <- paste0('https://records.nhl.com/site/api/franchise-', record_type, '-records')
  endpoint <- GET(URL) %>%
    content("text", encoding = 'UTF-8') %>%
    fromJSON(flatten = T) 
   df1 <- endpoint$data
  
  #if else statement to get team by name or ID
  ifelse(is.na(team_id) & is.na(team_name), franchise_records <- df1, 
         ifelse(is.na(team_id) & is.na(team_name) == FALSE, franchise_records <- filter(df1, franchiseName == team_name),
                ifelse(is.na(team_id) == FALSE & is.na(team_name), franchise_records <- filter(df1, franchiseId == team_id), 
                       franchise_records <- filter(df1, franchiseId == team_id | franchiseName == team_name)                                                         ))                                                         )
  return(franchise_records)
}

get_franchise_records(record_type = 'skater', team_id = 2)[1:6]
```

## Franchise Details Function
```{r franchise detail function}
get_franchise_detail <- function(team_id = NA){
  
  #connect to API drilling down if needed
  URL1 <- paste0('https://records.nhl.com/site/api/franchise-detail')
  URL2 <- paste0('https://records.nhl.com/site/api/franchise-detail?cayenneExp=mostRecentTeamId=', as.character(team_id))
  
  ifelse(is.na(team_id), URL <- URL1, URL <- URL2)

  endpoint <- GET(URL) %>%
    content("text", encoding = 'UTF-8') %>%
    fromJSON(flatten = T) 
  
  return(endpoint$data)
}

get_franchise_detail(10)
```

## Franchise Stats and Rankings Function
```{r get stats}
get_team_stats <- function(team_id = 'All', team_name = NA){

if (team_id == 'All'){
#teams info to show where individual team info is located
get_endpoint_all <- GET('https://statsapi.web.nhl.com/api/v1/teams/') %>%
    content("text", encoding = 'UTF-8') %>%
    fromJSON(flatten = T) 

#use URL pulled from get_endpoint_all to grab data from all teams and then combine
URL_combine <- paste0('https://statsapi.web.nhl.com/api/v1/teams/', as.character(get_endpoint_all$teams$id), '/stats')
#get all team info
team_stats_combined_single_season <- team_stats_combined_rankings <- vector("list", length(URL_combine))

  for (i in 1:length(URL_combine)){
    team_stat <- GET(URL_combine[i]) %>%
      content("text", encoding = 'UTF-8') %>%
      fromJSON(flatten = T) 
   team_stats_combined_single_season[[i]] <- team_stat$stats$splits[[1]]
   team_stats_combined_rankings[[i]] <- team_stat$stats$splits[[2]]
  }
team_stats_combined_single_season <- bind_rows(team_stats_combined_single_season)
team_stats_combined_rankings <- bind_rows(team_stats_combined_rankings)

final <- list(single_season = team_stats_combined_single_season, rankings = team_stats_combined_rankings)
if (is.na(team_name)) return(final)
  else return(list(single_season = filter(team_stats_combined_single_season, team.name == team_name), rankings = filter(team_stats_combined_rankings, team.name == team_name)))

} else {
#if specific team wanted
  URL <- paste0('https://statsapi.web.nhl.com/api/v1/teams/', as.character(team_id), '/stats')
  get_endpoint <- GET(URL) %>%
      content("text", encoding = 'UTF-8') %>%
      fromJSON(flatten = T) 
  single_season <- get_endpoint$stats$splits[[1]] %>% 
    select(team.id, team.name, everything())
  rankings <- get_endpoint$stats$splits[[2]] %>% 
    select(team.id, team.name, everything())
  
  final <- list(single_season = single_season, rankings = rankings)
  return(final)
}
}

#get_team_stats()
#get_team_stats(team_id = 1)
get_team_stats(team_name = 'New York Islanders')
```

## Wrapper Function
All functions have team_id and team_name as arguments to obtain desired record. 
```{r wrapper}
get_endpoint <- function(endpoint = c('franchise', 'team totals', 'records', 'details', 'stats'), ...){
  ifelse(endpoint == 'franchise', return(get_franchise(...)),
      ifelse(endpoint == 'team totals', return(get_franchise_team_totals(...)),
           ifelse(endpoint == 'records', return(get_franchise_records(...)),
                  ifelse(endpoint == 'details', return(get_franchise_detail(...)),
                         ifelse(endpoint == 'stats', return(get_team_stats(...)))))))
}

get_endpoint('team totals', team_id = 7)
```


# Exploratory Data Analysis
Now I can easily pull data from several of the API endpoints. I will now perform an Exploratory Data Analysis. 

## Contingency Tables 
### One-way Contingency Table
```{r}
details <- get_endpoint('details')
t <- table(details$active)
kable(t, col.names = c('Team Active?', 'Count'), caption = 'Count of Active Teams')
```
### Two-way contingency Table
```{r}
activity_year <- details %>%
  separate(firstSeasonId, into = 'first_year', sep = 4) #keep first year for readability
t2 <- table(activity_year$first_year, activity_year$active)
kable(t2, caption = 'Activity of Teams by First Active Year')
```
## Visuals
### Line Chart Visualizing Contigency Table from Above
```{r}
t3 <- data.frame(t2) %>% #make freguency table a data frame
  rename(Year = Var1, Active = Var2)

ggplot(t3, aes(x = Year, y = Freq)) +
  geom_line(aes(group = Active, color = Active), size = 2) +
  labs(y = 'Count of Teams', title = 'Count of Active and Inactive Teams by YFirst Active Year') +
  scale_color_discrete(name = NULL, labels = c('Inactive', 'Active')) +
  theme(axis.text.x = element_text(angle = -40))
```
  
Above we see the number of currently active and inactive teams by the year they were first active. We see all teams in this list that are inactive were first active before 1970. We have the most teams that were first active in 1967.  
  
### Bar Plot exploring Team Wins
```{r}
team_totals <- get_endpoint('team totals')

team_wins <- team_totals %>%
  group_by(teamName) %>%
  summarize_at(c('wins', 'homeWins', 'roadWins'), sum, na.rm = T) %>% #add up rows by team
  arrange(desc(wins)) %>% #sort by number of wins
  head(25) %>% #only keep top 25
  ggplot(aes(reorder(teamName, wins)))
team_wins + geom_col(aes(y = wins, fill = 'darkred')) + #fill show go in aes layer so legend shows up 
  geom_col(aes(y = homeWins, fill = 'darkblue')) + #add layer to show proportion of wins that are home
  coord_flip() + 
  theme_minimal() +
  theme(legend.position = c(0.9,0.2)) +
  labs(x = 'Team Name', y = 'Total Wins', title = 'The 25 Teams with Greatest Number of Wins') +
  scale_fill_manual(name = 'Win Type', labels = c('Home', 'Away'), values = c('darkblue','darkred'))

```
  
Here we have a bar plot of the total wins by team broken into home wins and away wins. Only the top 25 teams are shown. The Montreal Canadiens have the most total wins at nearly 4,000.  
  
### Scatter Plot of Fewest Goals by Power Play Percentage with Text Labels for each Team ID
```{r}
season_records <- get_endpoint('records', record_type = 'season') %>%
  rename(team.id = id)
stats <- get_endpoint('stats')[[1]] #extract first element of this list

#combine records and stats
records_stats <- left_join(season_records, stats, by = 'team.id') %>%
  select(team.id, franchiseName, everything(), -team.name)
#str(records_stats)

#create scatter plot
ggplot(records_stats, aes(fewestGoals, stat.powerPlayPercentage)) +
  geom_text(aes(label = team.id), na.rm = T) +
  labs(x = 'Fewest Goals', y = 'Power Play Percentage', title = 'Record for Fewest Goals by Power Play Percentage for Each Team ID') +
  theme_minimal()

```
  
The graph above shows the power play percentage by fewest goals in a season for each team. The points are labeled with the team ID for reference. We do not see much of a relationship, but we do notice team 54 has the largest number by far for fewest goals in a season.  
  
### Box plot for Number of Road Loss Streaks by Starting Month
```{r}
season_records %>% 
  select(starts_with('roadLoss')) %>% #select relevant data
  separate(roadLossStreakDates, into = 'start_month', sep = 3) %>% #separate column so I just have start month
  filter(is.na(start_month) == F) %>%
  mutate(month_numeric = ifelse(start_month == 'Jan', 13, 
                                ifelse(start_month == 'Feb', 14, match(start_month, month.abb)))) %>% #put Jan and Feb on the end so we are looking at the correct ordering of one season
  ggplot(aes(reorder(start_month, month_numeric), roadLossStreak)) +
    geom_boxplot() +
  labs(x = 'Start Month', y = 'Road Loss Streak', title = 'Road Loss Streaks by Start Month')

```

Above we have a box plot for the road loss streaks by start month ordered by the season months from October to February. October and December both have two outliers.  
  
### Histogram of the Distribution of the Proportion of losses per game for Goalies
```{r}
get_endpoint('records', record_type = 'goalie') %>%
  filter(gamesPlayed > 10) %>% #to cut down on those who played a just a few games and thus had extreme proportions
  transmute(losses_per_game = losses/gamesPlayed) %>% #create and keep only the new variable
  ggplot(aes(losses_per_game)) +
    geom_histogram(aes(y = ..density..), bins = 25, fill = 'darkblue') +
    geom_density(size = 2) +
    labs(x = 'Proportion of Games Lost', y = 'Density', title = 'Distribution of the Proportion of Games Lost for Goalies who Played >10 Games')
```

Above is the distribution of the proportion of games lost by individual goalies. Those who played less than 10 games were removed to avoid extreme proportions. A density curve is overlayed on the histogram. The mean proportion appears to be around 0.3, and the data is more or less normally distributed, though it has a slight skew to the right.  
  
### Scatter plot showing median penalty minutes by median rookie games
```{r}
skater <- get_endpoint('records', record_type = 'skater')
skater %>%
  group_by(franchiseName) %>%
  summarise(median_penalty_minutes = median(penaltyMinutes, na.rm = T), median_rookie_games = median(rookieGamesPlayed, na.rm = T)) %>%
  ggplot(aes(median_rookie_games, median_penalty_minutes)) +
    geom_point(size = 2) +
    geom_smooth(method = lm) +
    labs(x = 'Median Rookie Games', y = 'Median Penalty Minutes', title = 'Median Rookie Games by Median Penalty Minutes for each Team')
```

Finally, we have a scatter plot showing the median rookie games by the median penalty minutes for skaters, grouped by team. We see a slight positive trend here, as seen by the linear regression line. 
